Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-06-05 _tex/paper2.tex
Built site for gh-pages

To address these issues, this study extends Thurstone's general form
through a systematic and integrated approach that combines causal and
Bayesian inference methods. In addition to potentially enhancing
measurement reliability and validity, and improving statistical accuracy
in hypothesis testing, this approach offers two key advantages. First,
it clarifies the interactions among all actors and processes involved in
CJ experiments. Second, it shifts the current comparative data analysis
paradigm from passively accepting the BTL model assumptions to actively
testing whether those assumptions fit the data under analysis.

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's general
form to address these challenges. The extension integrates core
theoretical principles alongside key CJ experimental design features,
such as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model to analyze dichotomous
pairwise comparison data. Finally, Section~\ref{sec-discussion}
discusses the findings, explores avenues for future research, and detail
the challenges for future researchers.

@@

Thurstone noted from the outset that his general formulation, described
in Section~\ref{sec-thurstone_theory}, led to a \emph{trait scaling
problem}. Specifically, the model required estimating more ``unknown''
parameters than the number of available pairwise comparisons
\citep{Thurstone_1927b}. For instance, in a CJ experiment with five
texts, the general form would require estimating \(20\) parameters: five
modal discriminal processes, five discriminal dispersions, and \(10\)
correlations--one per comparison (see Table~\ref{tbl-thurstone_cases}).
However, a single judge could only provide \({5 \choose 2} = 10\) unique
comparisons, an insufficient data set to estimate the required
parameters.

@@

Notably, while assuming equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that this
assumption may fail to capture the complexity of some traits or account
for heterogeneous stimuli, such as handwritten texts or English
compositions
\citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022}.
Indeed, the presence of the so-called \emph{misfit} texts may already
signal these limitations, as these are texts that elicit more judgment
discrepancies than others
\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018},
possibly due to larger discriminal dispersions or because they are
genuine outliers--meaning, texts with distinctive characteristics that
deviate markedly from the rest of the sample \citep{Grubbs_1969}. In
either case, the BTL model neither accounts for these anomalies nor
provides tools for addressing them, apart from excluding the
``problematic'' texts from analysis.

@@

Despite these notable differences in the distribution of the discriminal
difference under various correlational assumptions, in practice,
experimental designs often adopt the assumption of no correlation
between stimuli based on an old theoretical justification. Specifically,
\citet{Thurstone_1927b} argued that stimuli could be treated as
uncorrelated because judges' biases--arising from two opposing and
equally weighted effects occurring during the pairwise
comparisons--would cancel each other out. This idea was later formalized
by \citet{Andrich_1978}, who provided a mathematical demonstration of
this cancellation using the BTL model under the assumption of
discriminal processes with additive biases. However, evidence from the
CJ literature indicates that the assumption of zero correlation does not
hold in practice in at least two scenarios: when intricate aspects of
multidimensional, complex traits or heterogeneous stimuli influence
judges' perceptions or when additional hierarchical structures are
relevant to the stimuli.

In the first scenario, research on text quality suggests that when
judges evaluate multidimensional, complex traits or heterogeneous
stimuli, they often rely on a variety of intricate stimulus
characteristics to inform their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
characteristics, regardless of their relevance, are unlikely to be
equally weighted or to oppose one another across comparisons. As a
result, they may exert a disproportionate influence on judges'
perceptions, generating biases that persist rather than cancel out. For
example, this could occur when a judge assessing the argumentative
quality of a text may place disproportionate emphasis on handwriting
clarity, thereby favoring neatly written texts despite their weaker
arguments. Moreover, because the discriminal process of stimuli becomes
an observable outcome only through the judges' perceptions, these biases
could introduce dependencies between the stimuli
\citep{vanderLinden_et_al_2017_II}. Evidence from traditional marking
contexts confirms a similar phenomenon \citep{Morin_et_al_2018}, and
additional studies
\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a}
support the presence of these biases in CJ--reinforcing the argument
that the factors influencing pairwise comparisons do not always cancel
each other out.

@@

Thus, erroneously assuming zero correlation between stimuli can also
lead to significant statistical and measurement issues. Specifically,
neglecting judges' biases or relevant hierarchical structures can create
dimensional mismatches in the model, leading to the over- or
underestimation of trait reliability
\citep{Ackerman_1989, Hoyle_et_al_2023} and even biases
\citep{Wu_et_al_2022}. These inaccuracies can result in spurious
conclusions about the discriminal differences \citep{McElreath_2020}
and, by extension, the underlying discriminal processes of the stimuli.
This issue is illustrated in Figure~\ref{fig-correlation} when the
discriminal difference distribution of the BTL scores follows the thick
continuous line \((\rho = 0)\), while the \emph{true} discriminal
difference follows any discontinuous line where \(\rho \neq 0\).

Finally, as discussed in the previous section, removing \emph{misfit}
judges--that is, judges whose assessments deviate markedly from the
shared consensus
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
and considered outliers under the BTL model
\citep{Grubbs_1969, Wu_et_al_2022}--risk discarding valuable information
and introducing bias into trait estimates. The direction and magnitude
of these biases remain unpredictable, as they depend on which judges
researchers exclude from the analysis
\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.

@@

The \emph{structural approach} to causal inference provides a formal
framework for identifying causes and estimating their effects using
data. The approach uses structural causal models (SCMs) and directed
acyclic graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
formally and graphically represent the assumed causal structure of a
system, such as the one found in CJ experiments. Essentially, SCMs and
DAGs function as \emph{conceptual models} on which identification
analysis rests. \emph{Identification analysis} determines whether an
estimator can accurately compute an estimand based solely on its
(causal) assumptions, regardless of random variability
\citep{Schuessler_et_al_2023}. Here, \emph{estimands} represent the
specific quantities researchers aim to determine (i.e., a parameter)
\citep{Everitt_et_al_2010}. \emph{Estimators} denote the methods or
functions that transform data into an estimate (e.g., a statistical
model), while \emph{estimates} are the numerical values approximating
the estimand \citep{Neal_2020, Everitt_et_al_2010}.

A motivating example that will appear throughout this document clarifies
these concepts. In this example, researchers aim to answer the question:
``To what extent do different teaching methods influence students'
ability to produce high-quality written texts?'' To investigate this, a
researcher designs a CJ experiment by randomly assigning students
(individuals) to two groups, each receiving a different teaching method.
Judges then compare pairs of students' written texts (stimuli) to
produce a dichotomous outcome reflecting the relative quality of each
text (trait). Based on this setup, researchers can reformulate the
research question as the estimand: ``\emph{On average}, is there a
difference in the ability to produce high-quality written texts between
the two groups of students?''.

@@

Thus, this section addresses the issues identified in
Section~\ref{sec-theory-issues} by extending Thurstone's general form
using the structural approach to causal inference. Specifically, it
combines the core theoretical principles outlined in
Section~\ref{sec-thurstone_theory} with key assessment design features
relevant to CJ experiments, such as the selection of judges, stimuli,
and comparisons. In addition to improving statistical accuracy and
strengthening measurement reliability and validity, the approach offers
two key advantages. First, it clarifies the interactions among all
actors and processes involved in CJ experiments. Second, it shifts the
current comparative data analysis paradigm from passively accepting the
model assumptions to actively testing whether those assumptions fit the
data under analysis.

@@

SCM~\ref{fig-cj03_scm} presents the first theoretical principles
embedded in the conceptual-population model, which evaluates the impact
of different teaching methods on students' writing ability. This SCM
outlines the relationship between the conceptual-population outcome
\((O^{cp}_{iahbjk})\) and several related variables. The subscripts
\(i\) and \(h\) identify the students who authored the texts (i.e., the
individuals). The indices \(a\) and \(b\) represent the texts under
comparison (i.e., the stimuli). The index \(j\) indicates the judge
conducting the comparison, while the index \(k\) accounts for
experimental conditions where a judge compares the same pair of stimuli
multiple times, i.e., a \emph{repeated measures designs}
\citep[pp.~366-376]{Lawson_2015}. Thus, the indexing system supports
comparisons between different texts written by the same student
\((i = h;\) \(a \neq b)\) and between texts written by distinct students
\((i \neq h;\) where \(a = b\) is permitted\()\), each compared once or
repeatedly by all judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K;\)
where \(n_{J}>1\) and \(n_{K}\geq1)\). However, it excludes cases where
a judge compares a student's text to itself, whether once or multiple
times \((i = h;\) \(a = b;\) \(j = 1,\dots,n_{J};\)
\(k = 1,\dots,n_{K};\) where \(n_{J}>1\) and \(n_{K}\geq1)\), as such
comparison lacks practical relevance within the CJ framework. Here,
\(n_{J}\) indicates the total number of judges, and \(n_{K}\) denotes
the number of repeated judgments each judge performs.

@@

In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
texts' discriminal processes \((T_{ia}, T_{hb})\) and their discriminal
difference \((D_{iahbjk})\) (see Section~\ref{sec-thurstone_theory}).
Additionally, the SCM incorporates a key design feature of CJ
experiments: the judges' biases \((B_{kj})\). This extension builds on
the arguments presented in Section~\ref{sec-theory-issue1b}, contending
that the discriminal difference becomes an observable outcome only
through judges' perceptions. Given that such perceptions may be
imperfect--and that each judge may carry some degree of bias
\citep[see][]{Pollitt_et_al_2003, vanDaal_et_al_2016}--it is reasonable
that judges' perceptions (bias) should be treated as an integral
component of the CJ system from the outset, as this leads to a more
accurate representation of the data-generating process underlying the
pairwise comparisons. This model defines the preliminary set of
endogenous variables,
\(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\), and the
preliminary set of structural equations, \(F = { f_{O}, f_{D} }\), which
capture the non-parametric dependencies among these variables.

@@

However, due to concerns about the practical feasibility of the
comparison task {[}\citet{Boonen_et_al_2020}; {]}, CJ experiments rarely
implement an exhaustive pairings of sampled judges, stimuli, and
individuals. Thus, a realistic scenario must account for the fact that
judges typically compare only a subset of stimuli authored by a sample
of individuals.

@@

In summary, the SCM~\ref{fig-cj15_scm} and DAG~\ref{fig-cj15_dag} extend
Thurstone's general form to address several limitations of the BTL
model. These extensions account for judge biases (see
Section~\ref{sec-theory-theoretical_P1}), reflect the hierarchical
structure of stimuli and incorporate measurement error in trait
estimation and hypothesis testing (see
Section~\ref{sec-theory-theoretical_P3}), and clarify the role of the
sample and comparison mechanisms in CJ experiments (see
Section~\ref{sec-theory-theoretical_SC}). However, they do not resolve
concerns about the assumption of equal dispersions among stimuli
discussed in Section~\ref{sec-theory-issue1a}. Since this concern
relates to the statistical assumption underlying the distribution of the
discriminal process, we develop a formal statistical model to address it
in the next section.

@@

Finally, we use \emph{Bayesian inference methods} to convert the
statistical model \ref{fig-cj16_stat} into a practical statistical tool
for analyzing paired comparison data. Bayesian inference offers three
main advantages in this context. First, it handles complex and
overparameterized models, where the number of parameters exceeds the
number of observations \citep{Baker_1998, Kim_et_al_1999}. This feature
is essential for our implementation, as the proposed model is indeed
overparameterized. Second, it incorporates prior information to
constrain parameter estimates within plausible bounds, thereby
mitigating estimation issues like non-convergence or improper solutions
that often affect frequentist methods
\citep{Martin_et_al_1975, Seaman_et_al_2011}. We primarily use this
capability to define the error distribution and set the scale of latent
variables \citep{Depaoli_2014}. Third, Bayesian inference supports
robust inferences from small samples, where the asymptotic properties
underlying frequentist methods are less reliable
\citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}. This
feature is particularly relevant in CJ experiments, as researchers often
collect large volumes of paired comparisons but work with relatively
small samples of judges, stimuli, and individuals to test hypotheses.

@@

Among these, Case V remains the most widely used model in empirical CJ
research, mainly due to the widespread adoption of the BTL model. The
BTL model mirrors the core assumptions of Case V--namely, equal
discriminal dispersions and zero correlation among stimuli' discriminal
processes--but replaces the processes' normal distribution with the more
mathematically tractable logistic distribution
\citep{Andrich_1978, Bramley_2008}. Notably, although these assumptions
simplify the trait measurement model, they may fail to capture the
complexity of some traits or account for heterogeneous stimuli
\citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022},
potentially leading to unreliable and inaccurate trait estimates
\citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023}.

Moreover, because Thurstone's original goal was to produce a ``coarse
scaling'' of traits and allocate stimuli along this continuum
\citep[pp.~269]{Thurstone_1927a}, his theory offered no guidance on how
to use trait estimates for statistical inference. To address this gap,
researchers in the CJ tradition have often separated trait estimation
from hypothesis testing, relying on point estimates of the traits--such
as BTL scores or their transformations--for inference. Notably, while
this approach simplifies analysis, it can also introduce bias and
compromise the reliability of the resulting inferences
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

Thus, to address these limitations of Thurstone's Case V and the BTL
model, this study extends Thurstone's general form using a systematic,
integrated approach that combines causal and Bayesian inference methods.
The approach begins by formulating a conceptual model represented by a
Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG). This
model integrates Thurstone's core theoretical principles, such as the
discriminal processes of stimuli, alongside key CJ experimental design
features, including judges' bias, sampling procedures, and comparison
mechanisms, thereby disentangling the causal processes in the CJ system.

The study then translates the SCM into a bespoke statistical model that
allows researchers to analyze CJ data when violations to the assumptions
of equal dispersion and zero correlation occur (see
Section~\ref{sec-theory-issue1a} and Section~\ref{sec-theory-issue1b})
and when statistical inference is necessary (refer to
Section~\ref{sec-theory-issue2}). In particular, this model accounts for
judge biases (see Section~\ref{sec-theory-theoretical_P1}), captures the
hierarchical structure of stimuli, incorporates measurement error into
the hypothesis testing process (refer to
Section~\ref{sec-theory-theoretical_P3}), and accommodates heterogeneity
in discriminal dispersions (see Section~\ref{sec-statistical}). By
addressing all these issues, these methodological innovations have the
potential to enhance the reliability and validity of trait measurement
\citep{Perron_et_al_2015} and improve the accuracy of statistical
inferences.

Beyond these potential benefits, the approach offers two additional
advantages. First, it clarifies the roles and interactions of all actors
and processes involved in CJ experiments. Second, it shifts the analytic
paradigm from passively accepting the assumptions of the BTL model to
actively testing their fit with observed data. Together, these
advantages establish a principled framework for evaluating best
practices in CJ experimental designs, one that better aligns with the
demands of contemporary CJ assessment contexts \citep{Kelly_et_al_2022},
offering new insights into existing research and opening promising
avenues for future inquiry.

\subsection{Further research}\label{sec-discussion_RA}

Four current and new research avenues deserve particular attention due
to their direct impact on the reliability and validity of trait
estimates and the accuracy of statistical inferences. The following
sections outline these avenues and explain how our model provides a
rigorous research framework to investigate them.

@@

Although sampling and comparison mechanisms are central to modern CJ
experimental design, it is striking that most of the CJ literature has
examined them only within a limited scope. Researchers have primarily
investigated the effects of adaptive comparative judgment (ACJ) designs
on trait reliability
\citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015, Verhavert_et_al_2022, Mikhailiuk_et_al_2021, Gray_et_al_2024}
or proposed practical guidelines for the number of comparisons judges
should make \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}.
Although these studies offer valuable insights, we argue that they have
overlooked the broader role these mechanisms play within the CJ system.
Moreover, we suggest that this oversight largely stems from a lack of
conceptual clarity about how the mechanisms function within the system.
To address this gap, this study decided to integrate these mechanisms
into CJ's conceptual model.

The explicit integration of the sampling and comparison mechanisms
offers a new perspective on how these mechanisms shape the CJ process.
In particular, it clarifies their role in the data-generating process by
framing them as sources of missing data--that is, as mechanisms that
determine which observations are missing from the final data sample.
This new perspective encourages the application of Little and Rubin's
principled missing data framework \citeyearpar{Little_et_al_2020},
enabling a more rigorous evaluation of existing claims about these
mechanisms, their influence on CJ outcomes, and their implications for
designing and assessing more complex experimental setups.

>>>>

To address these issues, this study extends Thurstone's general form
through a systematic and integrated approach that combines causal and
Bayesian inference methods. In addition to potentially enhancing
measurement reliability and validity, and improving statistical accuracy
in hypothesis testing, this approach offers two key advantages. First,
it clarifies the interactions among all actors and processes involved in
CJ assessments. Second, it shifts the current comparative data analysis
paradigm from passively accepting the BTL model assumptions to actively
testing whether those assumptions fit the data under analysis.

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's general
form to address these challenges. The extension integrates core
theoretical principles alongside key CJ assessment design features, such
as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model to analyze dichotomous
pairwise comparison data. Finally, Section~\ref{sec-discussion}
discusses the findings, explores avenues for future research, and detail
the challenges for future researchers.

@@

Thurstone noted from the outset that his general formulation, described
in Section~\ref{sec-thurstone_theory}, led to a \emph{trait scaling
problem}. Specifically, the model required estimating more ``unknown''
parameters than the number of available pairwise comparisons
\citep{Thurstone_1927b}. For instance, in a CJ assessment with five
texts, the general form would require estimating \(20\) parameters: five
modal discriminal processes, five discriminal dispersions, and \(10\)
correlations--one per comparison (see Table~\ref{tbl-thurstone_cases}).
However, a single judge could only provide \({5 \choose 2} = 10\) unique
comparisons, an insufficient data set to estimate the required
parameters.

@@

Notably, while assuming equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that this
assumption may fail accommodate for heterogeneous stimuli, such as
handwritten texts or English compositions
\citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022}.
Indeed, the presence of the so-called \emph{misfit texts}--texts that
elicit more judgment discrepancies than others
\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018}--may
signal this limitation, as such discrepancies can arise from larger
discriminal dispersions due to stimulus heterogeneity or because the
texts are genuine outliers--that is, texts with distinctive
characteristics that deviate markedly from the rest of the sample
\citep{Grubbs_1969}. In either case, the BTL model's assumptions prevent
it from adequately accounting for or addressing these anomalies, leaving
exclusion of the ``problematic'' texts as the primary remedy.

@@

Despite these notable differences in the distribution of the discriminal
difference under various correlational assumptions, in practice,
assessment designs often adopt the assumption of no correlation between
stimuli based on an old theoretical justification. Specifically,
\citet{Thurstone_1927b} argued that stimuli could be treated as
uncorrelated because judges' biases--arising from two opposing and
equally weighted effects occurring during the pairwise
comparisons--would cancel each other out. This idea was later formalized
by \citet{Andrich_1978}, who provided a mathematical demonstration of
this cancellation using the BTL model under the assumption of
discriminal processes with additive biases. However, evidence from the
CJ literature indicates that the assumption of zero correlation does not
hold in practice in at least two scenarios: when intricate aspects of
multidimensional, complex traits or heterogeneous stimuli influence
judges' perceptions or when additional hierarchical structures are
relevant to the stimuli.

In the first scenario, research on text quality suggests that when
judges evaluate complex, multidimensional traits or heterogeneous
stimuli, they often rely on a variety of intricate stimulus
characteristics to inform their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022}.
Regardless of their relevance, these characteristics may not receive
equal weight or consistently oppose one another across comparisons. As a
result, they may exert a disproportionate influence on judges'
perceptions, generating biases that persist rather than cancel out. For
example, this could occur when a judge assessing the argumentative
quality of a text may place disproportionate emphasis on handwriting
clarity, thereby favoring neatly written texts despite their weaker
arguments. Moreover, because the discriminal process of stimuli becomes
an observable outcome only through the judges' perceptions, these biases
could introduce dependencies between the stimuli
\citep{vanderLinden_et_al_2017_II}. While direct evidence for this exact
scenario is limited, existing studies document the presence of judge
bias in CJ \citep[\citet{vanDaal_et_al_2016},
\citet{Bartholomew_et_al_2020a}]{Pollitt_et_al_2003}, reinforcing the
argument that the factors influencing pairwise comparisons do not always
cancel each other out.

@@

Thus, erroneously assuming zero correlation between stimuli can also
lead to significant statistical and measurement issues. Specifically,
neglecting judges' biases or relevant hierarchical structures can create
dimensional mismatches in the model, leading to the over- or
underestimation of trait reliability
\citep{Ackerman_1989, Hoyle_et_al_2023} and even biases
\citep{Wu_et_al_2022}. This issue is illustrated in
Figure~\ref{fig-correlation} when the discriminal difference
distribution of the BTL scores follows the thick continuous line
\((\rho = 0)\), while the \emph{true} discriminal difference follows any
discontinuous line where \(\rho \neq 0\). Moreover, these inaccuracies
can result in spurious conclusions about the discriminal differences
\citep{McElreath_2020} and, by extension, the underlying discriminal
processes of the stimuli--potentially causing the model to incorrectly
classify stimuli or judges as \emph{misfits}.

Finally, similar to what we discussed in the previous section, removing
\emph{misfit} judges--that is, judges whose assessments deviate markedly
from the shared consensus
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
and may appear as outliers under the BTL model
\citep{Wu_et_al_2022}--risk discarding valuable information and
introducing bias into trait estimates. The direction and magnitude of
these biases remain unpredictable, as they depend on which judges
researchers exclude from the analysis
\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.

@@

The \emph{structural approach} to causal inference provides a formal
framework for identifying causes and estimating their effects using
data. The approach uses structural causal models (SCMs) and directed
acyclic graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
formally and graphically represent the assumed causal structure of a
system, such as the one found in CJ assessments. Essentially, SCMs and
DAGs function as \emph{conceptual models} on which identification
analysis rests. \emph{Identification analysis} determines whether an
estimator can accurately compute an estimand based solely on its
(causal) assumptions, regardless of random variability
\citep{Schuessler_et_al_2023}. Here, \emph{estimands} represent the
specific quantities researchers aim to determine (i.e., a parameter)
\citep{Everitt_et_al_2010}. \emph{Estimators} denote the methods or
functions that transform data into an estimate (e.g., a statistical
model), while \emph{estimates} are the numerical values approximating
the estimand \citep{Neal_2020, Everitt_et_al_2010}.

A motivating example that will appear throughout this document clarifies
these concepts. In this example, researchers aim to answer the question:
``To what extent do different teaching methods influence students'
ability to produce high-quality written texts?'' To investigate this, a
researcher designs a CJ assessment by randomly assigning students
(individuals) to two groups, each receiving a different teaching method.
Judges then compare pairs of students' written texts (stimuli) to
produce a dichotomous outcome reflecting the relative quality of each
text (trait). Based on this setup, researchers can reformulate the
research question as the estimand: ``\emph{On average}, is there a
difference in the ability to produce high-quality written texts between
the two groups of students?''.

@@

Thus, this section addresses the issues identified in
Section~\ref{sec-theory-issues} by extending Thurstone's general form
using the structural approach to causal inference. Specifically, it
combines the core theoretical principles outlined in
Section~\ref{sec-thurstone_theory} with key CJ assessment design
features, such as the selection of judges, stimuli, and comparisons. In
addition to improving statistical accuracy and strengthening measurement
reliability and validity, the approach offers two key advantages. First,
it clarifies the interactions among all actors and processes involved in
CJ assessments. Second, it shifts the current comparative data analysis
paradigm from passively accepting the model assumptions to actively
testing whether those assumptions fit the data under analysis.

@@

SCM~\ref{fig-cj03_scm} presents the first theoretical principles
embedded in the conceptual-population model, which evaluates the impact
of different teaching methods on students' writing ability. This SCM
outlines the relationship between the conceptual-population outcome
\((O^{cp}_{iahbjk})\) and several related variables. The subscripts
\(i\) and \(h\) identify the students who authored the texts (i.e., the
individuals). The indices \(a\) and \(b\) represent the texts under
comparison (i.e., the stimuli). The index \(j\) indicates the judge
conducting the comparison, while the index \(k\) accounts for assessment
conditions where a judge compares the same pair of stimuli multiple
times, i.e., a \emph{repeated measures designs}
\citep[pp.~366-376]{Lawson_2015}. Thus, the indexing system supports
comparisons between different texts written by the same student
\((i = h;\) \(a \neq b)\) and between texts written by distinct students
\((i \neq h;\) where \(a = b\) is permitted\()\), each compared once or
repeatedly by all judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K;\)
where \(n_{J}>1\) and \(n_{K}\geq1)\). However, it excludes cases where
a judge compares a student's text to itself, whether once or multiple
times \((i = h;\) \(a = b;\) \(j = 1,\dots,n_{J};\)
\(k = 1,\dots,n_{K};\) where \(n_{J}>1\) and \(n_{K}\geq1)\), as such
comparison lacks practical relevance within the CJ framework. Here,
\(n_{J}\) indicates the total number of judges, and \(n_{K}\) denotes
the number of repeated judgments each judge performs.

@@

In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
texts' discriminal processes \((T_{ia}, T_{hb})\) and their discriminal
difference \((D_{iahbjk})\) (see Section~\ref{sec-thurstone_theory}).
Additionally, the SCM incorporates a key CJ design feature: the judges'
biases \((B_{kj})\). This extension builds on the arguments presented in
Section~\ref{sec-theory-issue1b}, contending that the discriminal
difference becomes an observable outcome only through judges'
perceptions. Given that such perceptions may be imperfect--and that each
judge may carry some degree of bias
\citep[see][]{Pollitt_et_al_2003, vanDaal_et_al_2016}--it is reasonable
that judges' perceptions (bias) should be treated as an integral
component of the CJ system from the outset, as this leads to a more
accurate representation of the data-generating process underlying the
pairwise comparisons. This model defines the preliminary set of
endogenous variables,
\(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\), and the
preliminary set of structural equations, \(F = { f_{O}, f_{D} }\), which
capture the non-parametric dependencies among these variables.

@@

However, due to concerns about the practical feasibility of the
comparison task {[}\citet{Boonen_et_al_2020}; {]}, CJ assessments rarely
implement an exhaustive pairings of sampled judges, stimuli, and
individuals. Thus, a realistic scenario must account for the fact that
judges typically compare only a subset of stimuli authored by a sample
of individuals.

@@

In summary, the SCM~\ref{fig-cj15_scm} and DAG~\ref{fig-cj15_dag} extend
Thurstone's general form to address several limitations of the BTL
model. These extensions account for judge biases (see
Section~\ref{sec-theory-theoretical_P1}), reflect the hierarchical
structure of stimuli and incorporate measurement error in trait
estimation and hypothesis testing (see
Section~\ref{sec-theory-theoretical_P3}), and even clarify the role of
the sample and comparison mechanisms in CJ assessments (see
Section~\ref{sec-theory-theoretical_SC}). However, they do not resolve
concerns about the assumption of equal dispersions among stimuli
discussed in Section~\ref{sec-theory-issue1a}. Since this concern
relates to the statistical assumption underlying the distribution of the
discriminal process, we develop a formal statistical model to address it
in the next section.

@@

Finally, we use \emph{Bayesian inference methods} to convert the
statistical model \ref{fig-cj16_stat} into a practical statistical tool
for analyzing paired comparison data. Bayesian inference offers three
main advantages in this context. First, it handles complex and
overparameterized models, where the number of parameters exceeds the
number of observations \citep{Baker_1998, Kim_et_al_1999}. This feature
is essential for our implementation, as the proposed model is indeed
overparameterized. Second, it incorporates prior information to
constrain parameter estimates within plausible bounds, thereby
mitigating estimation issues like non-convergence or improper solutions
that often affect frequentist methods
\citep{Martin_et_al_1975, Seaman_et_al_2011}. We primarily use this
capability to define the error distribution and set the scale of latent
variables \citep{Depaoli_2014}. Third, Bayesian inference supports
robust inferences from small samples, where the asymptotic properties
underlying frequentist methods are less reliable
\citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}. This
feature is particularly relevant in CJ assessments, as researchers often
collect large volumes of paired comparisons but work with relatively
small samples of judges, stimuli, and individuals to test hypotheses.

@@

Among these, Case V remains the most widely used model in empirical CJ
research, mainly due to the widespread adoption of the BTL model. The
BTL model mirrors the core assumptions of Case V--namely, equal
discriminal dispersions and zero correlation among stimuli' discriminal
processes--but replaces the processes' normal distribution with the more
mathematically tractable logistic distribution
\citep{Andrich_1978, Bramley_2008}. Although this substitution has
minimal impact on trait estimation or model interpretation
\citep{vanderLinden_et_al_2017_I, McElreath_2021}, the simplifying
assumptions of the BTL model--and by extension, of Case V--may fail to
capture the complexity of some traits or account for heterogeneous
stimuli
\citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022},
potentially leading to unreliable and inaccurate trait estimates
\citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023}.

Moreover, because Thurstone's original goal was to produce a ``coarse
scaling'' of traits and allocate stimuli along this continuum
\citep[pp.~269]{Thurstone_1927a}, his theory offered no guidance on how
to use trait estimates for statistical inference. The CJ tradition has
attempted to address this gap by separating trait estimation from
hypothesis testing, relying on point estimates--such as BTL scores or
their transformations--for inference. Notably, while this approach
simplifies analysis, it can also introduce bias and compromise the
reliability of the resulting inferences
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

Thus, to address the limitations of Thurstone's Case V and the BTL
model, this study extends Thurstone's general form using a systematic,
integrated approach that combines causal and Bayesian inference methods.
The approach begins by formulating a conceptual model represented by a
Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG). This
model integrates Thurstone's core theoretical principles, such as the
discriminal processes of stimuli, alongside key CJ assessment design
features, including judges' bias, sampling procedures, and comparison
mechanisms, thereby disentangling the causal processes in the CJ system.

The approach then translates the SCM into a bespoke statistical model
that allows researchers to analyze CJ data when violations to the
assumptions of equal dispersion and zero correlation occur (see
Section~\ref{sec-theory-issue1a} and Section~\ref{sec-theory-issue1b})
and when statistical inference is necessary (refer to
Section~\ref{sec-theory-issue2}). In particular, this model accounts for
judge biases (see Section~\ref{sec-theory-theoretical_P1}), captures the
hierarchical structure of stimuli, incorporates measurement error into
the hypothesis testing process (refer to
Section~\ref{sec-theory-theoretical_P3}), and accommodates heterogeneity
in discriminal dispersions (see Section~\ref{sec-statistical}). By
addressing all these issues, these methodological innovations have the
potential to enhance the reliability and validity of trait measurement
in CJ \citep{Perron_et_al_2015}, while also improving the accuracy of
statistical inferences.

Beyond these potential benefits, the approach offers two additional
advantages. First, it clarifies the roles and interactions of all actors
and processes involved in CJ assessments. Second, it shifts the analytic
paradigm from passively accepting the assumptions of the BTL model to
actively testing their fit with observed data. Together, these
advantages establish a principled framework for evaluating best
practices in designing CJ assessments, one that better aligns with the
demands of contemporary CJ contexts \citep{Kelly_et_al_2022}, offering
new insights into existing research and opening promising avenues for
future inquiry.

\subsection{Future research directions based on this study's
approach}\label{sec-discussion_RA}

Among the many potential directions for future research--both current
and new--three avenues deserve particular attention due to their direct
impact on the reliability and validity of trait estimates, as well as on
the accuracy of statistical inferences. The following sections outline
these avenues and explain how our approach provides a rigorous research
framework to investigate them.

@@

Although sampling and comparison mechanisms are central to modern CJ
assessments, it is striking that most of the CJ literature has examined
them only within a limited scope. Researchers have primarily
investigated the effects of adaptive comparative judgment (ACJ) designs
on trait reliability
\citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015, Verhavert_et_al_2022, Mikhailiuk_et_al_2021, Gray_et_al_2024}
or proposed practical guidelines for the number of comparisons judges
should make \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}. While
these studies offer valuable insights, they often overlook the broader
role these mechanisms play within the CJ system. Moreover, as this
oversight likely stems from a broader lack of conceptual clarity about
how these mechanisms function within the system, this study formally
incorporates these mechanisms into CJ's conceptual model.

The explicit integration of the sampling and comparison mechanisms
offers a new perspective on how these mechanisms shape the CJ process.
In particular, it clarifies their role in the data-generating process by
framing them as sources of missing data--that is, as mechanisms that
determine which observations are missing from the final data sample.
This new perspective encourages the application of Little and Rubin's
principled missing data framework \citeyearpar{Little_et_al_2020},
enabling a more rigorous evaluation of existing claims about these
missing data mechanisms, their influence on CJ outcomes, and their
implications for designing and assessing more complex assessments
setups.
